Approximating the end of nested sampling

### 2023-10-25 paper/paper.tex
revising paper

%{{{ Preamble
\documentclass[usenatbib]{mnras}
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage{layouts}
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
\usepackage{fancyvrb}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{txfonts}
\usepackage[T1]{fontenc}
%}}}

\renewcommand\thesubsubsection{\Alph{subsubsection}.}
\newcommand{\nlive}{n_\mathrm{live}}
\newcommand{\Like}{\mathcal{L}}
\newcommand{\DKL}{\mathcal{D}_\mathrm{KL}}
\newcommand{\logLmax}{\log \Like_\mathrm{max}}

@@

\begin{abstract}
\end{abstract}

@@

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously functions as a probabilistic sampler, integrator and optimiser \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{sparse_reconstruction}. The core algorithm is unique for estimating volumes by \textit{counting}, which makes high-dimensional integration feasible. It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
\par
The order of magnitude runtime of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user. Currently, existing implementations of nested sampling \citep{multinest, polychord} either do not give an indication of runtime, or only provide crude measures of progress that do not directly correspond to the runtime. 
\par
This paper sets out a more principled manner of endpoint estimation for nested sampling at each intermediate stage, the key idea being to use the existing samples to predict the likelihood in the region we have yet to sample from. Outline of paper.
\begin{figure}
\begin{Verbatim}[frame=single, commandchars=\\\{\}]
\textcolor{red}{Predicted endpoint: 25054 +/- 242}
\textcolor{red}{Progress: [=================>########] 72%}
___________________
lives      |   500 |
phantoms   | 24310 |
posteriors | 18018 |
equals     |   245 |
-------------------
ncluster   =  1/1
ndead      =  18018
nposterior =  18018
nequals    =  249
nlike      =  4159049
<nlike>    =  491.04 (9.82 per slice)
log(Z)     =  -12.55 +/- 0.27
\end{Verbatim}
\caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}
\label{fig:polychord_output}
\end{figure}

\section{Background}
Let us begin with a brief description of the nested sampling algorithm to establish the necessary notation. For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
\begin{equation}\label{eq:evidence}
	\mathcal{Z} = \int \mathcal{L}(\theta)\ \pi(\theta)\ \mathrm{d}\theta
\end{equation}
while producing samples of the posterior distribution
\begin{equation}
	\mathcal{P}(\theta) = \frac{\mathcal{L}(\theta) \pi(\theta)}{\mathcal{Z}}.
\end{equation}
The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run \citep{dynamic_ns}. At each iteration, the point with the lowest likelihood is removed and added to a list of \textit{dead points}. A new point is then drawn from the prior, subject to the constraint that it must have a higher likelihood than the latest dead point. Repeating the procedure leads to the live points shrinking around peaks in the likelihood.
\par
The integral in \cref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
\begin{equation}
	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \mathcal{L}(X_{i-1}-X_{i+1}),
\end{equation}
where $X(\mathcal{L})$ is the fraction of the prior with a likelihood greater than $\mathcal{L}$. The prior volumes $X_i$ are unknown, but can be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
\begin{equation}\label{eq:X_dist}
	X_i = \prod_{k=1}^i t_k.
\end{equation}
The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
\begin{equation}\label{eq:t_dist}
	P(t_i) = \nlive t_i^{\nlive-1}, \quad \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
\end{equation}
The algorithm terminates when an user-specified condition is met; a popular choice is to terminate when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g. $10^{-3}$. The remaining live points are then killed off one by one without replacement and added to the evidence.
\par
Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$. For any given problem, the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{\nlive}$, so $\nlive$ sets the resolution.

\section{The anatomy of a nested sampling run}
The following section aims to make an inventory of the information available to us at an intermediate iteration $i^{*}$, which we shall eventually use to make endpoint predictions. We present an anatomy of the progression of a nested sampling run in terms of the prior volume compression, the log-likelihood increase, the inferred temperature schedule, and the dimensionality of the samples.

\subsection{Prior volume}
The key feature of nested sampling is that the sampling is controlled by prior volume compression. The task is to find the posterior typically lying in a tiny fraction of the prior volume, a total compression which is quantified by the average information gain, or \textit{Kullback-Leibler divergence}:
\begin{equation}\label{eq:DKL}
   \DKL = \int \mathcal{P}(\theta) \log \frac{\mathcal{P}(\theta)}{\pi(\theta)}\ \mathrm{d}\theta. 
\end{equation}
The bulk of the posterior lies within a prior volume $X = e^{-\DKL}$, which is the target compression. One gets there by iteratively taking steps of size $\Delta \log X_i = -1/n_i$, so that 
\begin{equation}
    \mathrm{E}(\log X_i) = -\sum_{k=1}^i \frac{1}{n_k}, \quad \mathrm{Var}(\log X_i) = \sum_{k=1}^i \frac{1}{n_k^2}.
\end{equation}
A constant step size in $\log X$ corresponds to a geometrically decreasing measure for the dead points (as shown in fig X), which is exactly needed to overcome the curse of dimensionality.
\par
Dead measure plot
\par
The same is not true for the live points, which are uniformly distributed in prior volume. As a result, the maximum live point is found at 
\begin{equation}\label{eq:Xmin}
	\mathrm{E}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{E}[\log X_*] - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i_*}{\nlive} - \log \nlive - \gamma,
\end{equation}
with variance
\begin{equation}
	\mathrm{Var}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{Var}[\log X_*] + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i_*}{\nlive^2} + \frac{\pi^2}{6}, 
\end{equation}
where the large $\nlive$ limit is taken for the approximation to the harmonic series, $\gamma$ being the Euler-Mascheroni constant.
\par
Hence the live points only get us a factor of $\log \nlive$ closer to the posterior bulk. In other words, it is not until we are around $\log \nlive$ away from $\log X = \DKL$ that the samples look anything like the posterior. One can see from \cref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
\par
Intuitively, it is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run.

\subsection{Log-likelihood}
It is also useful to observe the distribution of the live and dead points in log-likelihood. We choose to examine a representative case of the $d$-dimensional multivariate Gaussian:
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2
\end{equation}
with maximum point $\logLmax$ and lengthscale $\sigma$ to get an insight into the analytics. The likelihood monotonically increases towards $\logLmax$, with the posterior samples eventually concentrated around the bulk at
\begin{equation}
    \langle\log\Like\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
We can again find the size of each step in log-likelihood, as well as the expected location of the maximum live point. Define a likelihood normalised by the distance to the maximum 
\begin{equation}
    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
    \label{eq:normalised_likelihood}
\end{equation}
as a measure of how far a point is between the current likelihood and the maximum; $y=0$ corresponds to $\log\mathcal{L}_*$ and $y=1$ to $\log\mathcal{L}_\mathrm{max}$. At each iteration, $y$ increases by roughly
\begin{equation}
    \lim_{\nlive \to \infty}\Delta y \approx \frac{\mathrm{d} y}{\mathrm{d}\log X} \Delta \log X = \frac{2}{d \nlive}  
\end{equation}
in the large $\nlive$ limit. We can again sum the harmonic series to see that the maximum live point is expected to be at $y = 2\log \nlive/d$: a very small fraction in high dimensions, showing again that until the end the live points are far from the posterior bulk, and certainly nowhere near the maximum.
\par
The above also implies that the normalised distance between the highest and second highest live point is roughly $2/d$. Before reaching the posterior bulk, $\logLmax - \log\Like^{*} > d/2$, so the log-likelihood distance between the two points is larger than one, hence at least an order of magnitude apart in likelihood. It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.

\subsubsection*{Aside: nested sampling as a maximiser}
Previous literature \citep{Akrami_2010, Feroz_2011} has explored the potential for nested sampling to be used as a global maximiser, given its ability to handle multi-modalities. In particular, the latter authors emphasised that posterior samplers such as nested sampling find the bulk of the \textit{mass}, not the maximum of the distribution, but that this can be remedied by tightening the termination criterion. We now use the machinery we have developed to put this statement in more quantitative terms. 
\par
A more rigorous derivation (see Appendix) shows that the maximum live point has mean and variance
\begin{equation}\label{eq:ylivemax}
	\lim_{d,\nlive\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log \nlive}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}.
\end{equation}
Now let us take the dead point to be the termination point with likelihood $\log\Like_\mathrm{end}$ and prior volume $X_\mathrm{end}$, so that
\begin{equation}
	\epsilon = \frac{\int_0^{X_\mathrm{end}} \mathcal{L}\ \mathrm{d}X}{\int_0^\infty \mathcal{L}\ \mathrm{d}X}.
\end{equation}
Note that we have assumed that prior effects are negligible (so $1=\infty$), and that $\epsilon \ll 1$ so that the denominator is approximately the accumulated evidence. Computing this for \cref{eq:gaussian_logL}, we find the answer in terms of lower incomplete gamma functions
\begin{equation}
\epsilon = 1- \frac{\Gamma_{d/2}\left(X_\mathrm{end}^{2/d}/2\sigma^2\right)}{\Gamma(d/2)}.
\end{equation}
Taking the $X_\mathrm{end}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
\begin{equation}
    \lim_{X_\mathrm{end}\ll (\sqrt{2}\sigma)^d} \epsilon \approx \frac{X_\mathrm{end}}{(\sqrt{2}\sigma)^d \ \Gamma(1+\frac{d}{2})} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{end})^{\frac{d}{2}}}{\Gamma(1+\frac{d}{2})}.
\end{equation}
We thus have an expression relating $\mathcal{L}_\mathrm{end}$ at termination to the termination fraction $\epsilon$. This becomes yet more pleasing in the large $d$ limit, since $\epsilon^{2/d}\to 1$, we find via a Stirling approximation:
\begin{equation}
    \lim_{d\to\infty} \log\mathcal{L}_\mathrm{end} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e}.
\end{equation}
In the event that we keep $\epsilon$ in, we replace $\frac{d}{2e}\to \frac{d}{2e}\epsilon^{2/d}$, so we can of course battle the $\frac{d}{2e}$ term, but this becomes exponentially difficult in high dimensions.
\par
Putting this together, taking $\mathcal{L}_*$ in \cref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{end}$, and combining this with \cref{eq:ylivemax} we find
\begin{equation}
    \boxed{
        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
    },
\end{equation}
showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood. The final set of $n$ live points gets you $\log n/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.

\subsection{Temperature}
\subsubsection*{Motivations}
As shown in the previous section, midway through the run nearly all of the posterior mass is concentrated at a single point. However, this does not capture the \textit{structure} of the posterior that has been explored and all of the information it provides. 
\par
We have the potential to fix this because nested sampling is invariant to monotonic transformations; softening the likelihood via $L \to L^{\beta}$ brings the likelihoods closer together. Large $\beta$ and small $\beta$ both discard information, but there is an inbetween range of $\beta$ which is significant, because it causes the posterior mass to be concentrated at the current point; see plot. 
\par
At this point relevant to note the correspondence between Bayesian inference and statistical mechanics, from which the above transform is derived. If one takes $- \log \Like$ to be the energy $E_i$ etc, then $\beta$ is the inverse temperature $1/T$. Softening or hardening is equivalent to setting the temperature of the samples, which one can easily do in NS because it is athermal; get the posterior for $\Like^{\beta}$ trivially by reweighting.  
\par
Bayesian inference has a correspondence to statistical mechanics in that one can equate the parameters to microstates $i$, the negative log-lihood to the microstate energy $E_i$, and the prior to the density of states $g_i$. One can generalise Bayes' theorem to 
\begin{equation}
p(E_i) = \frac{g_i e^{-\beta E_i}}{Z(\beta)} \quad \leftrightarrow \quad \mathcal{P}_\beta(\theta) = \frac{\mathcal{L}^{\beta}(\theta)\pi(\theta)}{Z(\beta)}
\end{equation}
Clear similarity to thermodynamic integration, which uses temperature as a control parameter to construct intermediate distributions between the prior at $\beta = 0$ and posterior at $\beta = 1$. NS uses compression instead of temp, sidestepping difficulty of designing the annealing schedule. However, it still does something similar, starting with unweighted prior samples and ending with likelihood-weighted prior samples i.e. the posterior, during which $\beta$ presumably increases from 0 to 1. Inferring the temperature from above, one would hope to get another quantity that determines the progress of a run. 
\par
Important difference to annealing is that here we are working backwards; instead of choosing a temperature to sample at, we obtain the temperature from the samples after the fact. 
An useful discussion can be found in \citet{demons}, which we summarise here as a starting point. In contrast to annealing methods which attempt to sample from a series of canonical ensembles $\pi(\theta) \exp{-\beta \mathrm{E}(\theta)}$, nested sampling evolves a series of \textit{microcanonical} ensembles and computing the density of states by tracking the compression. 
\par
We now present several methods of obtaining the temperature that one can plausibly consider to be the current temperature of a nested sampling run.

\subsubsection{Microcanonical temperature}
As discussed in Skilling's original paper, the density of states, given by the negative slope of the log-likelihood curve, directly corresponds to a temperature;
\begin{equation}
    \beta^* = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like} \Bigg\vert_{\log \Like^{*}}.
\end{equation}
This is the microcanonical temperature $\partial S/\partial E$, where the volume entropy $S=\log X$ and the energy $E=-\log \Like$. Its value can be easily obtained via finite difference of the $\log \Like$ and $\log X$ intervals, albeit subject to an arbitrary window size for the differencing (in practice, we find that above a threshold of around 10 iterations the result is fairly insensitive to this choice).

\subsubsection{Canonical temperature}
Briefly considered by Habeck, it is the temperature of an ensemble whose average energy is the current energy, which one can obtain by inverting 
\begin{equation}
    \langle \log \Like \rangle_{\mathcal{P}_\beta} = \log \Like^{*}
\end{equation}

>>>>

%{{{ Preamble
\documentclass[usenatbib]{mnras}
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage{layouts}
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
\usepackage{fancyvrb}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage[T1]{fontenc}
%}}}

\renewcommand\thesubsubsection{\Alph{subsubsection}.}
\newcommand{\nlive}{n_i}
\newcommand{\Like}{\mathcal{L}}
\newcommand{\DKL}{\mathcal{D}_\mathrm{KL}}
\newcommand{\logLmax}{\log \Like_\mathrm{max}}

@@

\begin{abstract}
This paper develops a technique to estimate the runtime of nested sampling which works for any nested sampling implementation. At each iteration, the dimensionality of the current set of samples is inferred, then used to extrapolate the known likelihood profile using a Gaussian model function, which allows the remaining evidence and prior volume at termination to be calculated. The method finds the right order of magnitude at all times, obtaining the true endpoint within standard error generally around the halfway point, and converges to the true value by the end of the run.
\end{abstract}

@@

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously functions as a probabilistic sampler, integrator and optimiser \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{sparse_reconstruction}. The core algorithm is unique in its estimation of volumes by \textit{counting}, which makes high-dimensional integration feasible. It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
\par
The order of magnitude runtime of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user. Currently, existing implementations of nested sampling (e.g. \citealt{multinest, polychord, ultranest}) either do not give an indication of runtime, or only provide crude measures of progress that do not directly correspond to the runtime. 
\par
This paper sets out a principled manner of endpoint estimation for nested sampling at each intermediate stage (as shown in \cref{fig:polychord_output}), the key idea being to use the existing samples to predict the likelihood in the region we have yet to sample from. We begin with an overview of nested sampling in \cref{sec:background}, followed by an examination of the anatomy of a nested sampling run to establish key concepts for endpoint prediction. \cref{sec:endpoint} then outlines the methodology we use, including discussion and comparisons to previous attempts. Finally, \cref{sec:results} presents the results and discussions for toy and cosmological chains, before we conclude.
\begin{figure}
\begin{Verbatim}[frame=single, commandchars=\\\{\}]
\textcolor{red}{Predicted endpoint: 25054 +/- 242}
\textcolor{red}{Progress: [=================>########] 72%}
___________________
lives      |   500 |
phantoms   | 24310 |
posteriors | 18018 |
equals     |   245 |
-------------------
ncluster   =  1/1
ndead      =  18018
nposterior =  18018
nequals    =  249
nlike      =  4159049
<nlike>    =  491.04 (9.82 per slice)
log(Z)     =  -12.55 +/- 0.27
\end{Verbatim}
\caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}
\label{fig:polychord_output}
\end{figure}

\section{Background}\label{sec:background}
Let us begin with a brief description of the nested sampling algorithm to establish the necessary notation; for a more comprehensive treatment, we recommend Skilling's paper and the textbook \citet{sivia}. For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
\begin{equation}\label{eq:evidence}
	\mathcal{Z} = \int \mathcal{L}(\theta)\pi(\theta)\ \mathrm{d}\theta
\end{equation}
while producing samples of the posterior distribution
\begin{equation}
	\mathcal{P}(\theta) = \frac{\mathcal{L}(\theta) \pi(\theta)}{\mathcal{Z}}.
\end{equation}
The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run \citep{dynamic_ns}. At each iteration, the point with the lowest likelihood is removed and added to a list of \textit{dead points}. A new point is then drawn from the prior, subject to the constraint that it must have a higher likelihood than the latest dead point. Repeating the procedure leads to the live points shrinking around peaks in the likelihood.
\par
The integral in \eqref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
\begin{equation}
	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \mathcal{L}(X_{i-1}-X_{i+1}),
\end{equation}
where $X(\mathcal{L})$ is the fraction of the prior with a likelihood greater than $\mathcal{L}$. The prior volumes $X_i$ are unknown, but can be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
\begin{equation}\label{eq:X_dist}
	X_i = \prod_{k=1}^i t_k.
\end{equation}
The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
\begin{equation}\label{eq:t_dist}
	P(t_i) = \nlive t_i^{\nlive-1}
\end{equation}
\begin{equation}\label{eq:t_moments}
    \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
\end{equation}
The algorithm terminates when an user-specified condition is met; a popular choice is to terminate when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g. $10^{-3}$. The remaining live points are then killed off one by one without replacement and added to the evidence.
\par
Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$. For any given problem, the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{\nlive}$, so $\nlive$ sets the resolution of the algorithm.

\section{The anatomy of a nested sampling run}
The following section aims to make an inventory of the information available to us at an intermediate iteration $i^{*}$, which we shall eventually use to make endpoint predictions. We present an anatomy of the progression of a nested sampling run in terms of the prior volume compression, the log-likelihood increase, the inferred temperature, and the dimensionality of the samples.

\subsection{Prior volume}
The key feature of nested sampling is that the sampling is controlled by prior volume compression. The task is to find the posterior typically lying in a tiny fraction of the prior volume, a total compression which is quantified by the average information gain, or \textit{Kullback-Leibler divergence}:
\begin{equation}\label{eq:DKL}
   \DKL = \int \mathcal{P}(\theta) \log \frac{\mathcal{P}(\theta)}{\pi(\theta)}\ \mathrm{d}\theta. 
\end{equation}
The bulk of the posterior lies within a prior volume $X = e^{-\DKL}$, which is the target compression. One gets there by iteratively taking steps of size $\Delta \log X_i = -1/n_i$, so that when we add up the contribution of each step in \eqref{eq:t_moments} we get
\begin{equation}
    \mathrm{E}(\log X_i) = -\sum_{k=1}^i \frac{1}{n_k}, \quad \mathrm{Var}(\log X_i) = \sum_{k=1}^i \frac{1}{n_k^2}.
\end{equation}
A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points (as shown in \cref{fig:dead_measure}), which is exactly needed to overcome the curse of dimensionality.
\begin{figure*}
\begin{center}
    \includegraphics{figures/dead_measure_live.pdf}
\end{center}
\caption{The dead points of a nested sampling run, recursively zoomed in. Their density is constant in $\log X$, which is a geometrically constant measure $\propto \mathrm{d} V/V$, and hence scale invariant. The live points (larger dots) are uniform across the prior, plotted for comparison.}
\label{fig:dead_measure}
\end{figure*}
\par
The same is not true for the live points, which are uniformly distributed in prior volume. As a result, the maximum live point is found at 
\begin{equation}\label{eq:Xmin}
	\mathrm{E}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{E}(\log X_*) - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i_*}{\nlive} - \log \nlive - \gamma,
\end{equation}
with variance
\begin{equation}
	\mathrm{Var}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{Var}(\log X_*) + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i_*}{\nlive^2} + \frac{\pi^2}{6}, 
\end{equation}
where the large $\nlive$ limit is taken for the approximation to the harmonic series, $\gamma$ being the Euler-Mascheroni constant.
\par
The live points therefore only get us a factor of $\log \nlive$ closer to the posterior bulk. In other words, it is not until we are around $\log \nlive$ away from $\log X = -\DKL$ that the samples look anything like the posterior. One can see from \eqref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
Intuitively, it is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run.
\par
The above results are summarised in \cref{fig:logX_distribution}
\begin{figure*}
\begin{center}
    \includegraphics{figures/logX_distribution.pdf}
\end{center}
\caption{The distribution of the posterior mass in terms of $\log X$, the live points over the constrained prior and the smallest live point prior volume $\log X_\mathrm{min}^{\mathrm{live}}$ at an intermediate iteration $i_*$. For large values of $\DKL$ i.e. informative posteriors and/or large dimensionalities, the maximum live point is very far from the posterior bulk until the very end of the run. Note that the x-axis in this plot is $\log X$, so that the run proceeds from right to left to emphasise that the enclosed prior volume iteratively gets smaller. Plots of the sort from here onwards will be in terms of $-\log X$, where the run will more naturally proceed from left to right. }
\label{fig:logX_distribution}
\end{figure*}

\subsection{Log-likelihood}\label{sec:logL}
It is also useful to observe the distribution of the live and dead points in log-likelihood. We choose to examine a representative case of the $d$-dimensional multivariate Gaussian:
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2
\end{equation}
with maximum point $\logLmax$ and lengthscale $\sigma$ to get an insight into the analytics. The likelihood monotonically increases towards $\logLmax$, with the posterior samples eventually concentrated around the bulk at
\begin{equation}
    \langle\log\Like\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
We can again find the size of each step in log-likelihood, as well as the expected location of the maximum live point. Let us define a likelihood normalised by the distance to the maximum:
\begin{equation}
    y = \frac{\log\mathcal{L}-\log\mathcal{L}^*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}^*};
    \label{eq:normalised_likelihood}
\end{equation}
$y=0$ corresponds to the current point $\log\mathcal{L}^*$ and $y=1$ to the maximum $\log\mathcal{L}_\mathrm{max}$. At each iteration, $y$ increases by roughly
\begin{equation}
    \lim_{\nlive \to \infty}\Delta y \approx \frac{\mathrm{d} y}{\mathrm{d}\log X} \Delta \log X = \frac{2}{d \nlive}  
\end{equation}
in the large $\nlive$ limit. We can again sum the harmonic series to see that the maximum live point is expected to be at 
\begin{equation}
    y_{(\nlive)}^{\mathrm{live}} = \frac{2\log \nlive}{d},
\end{equation}
where we have used the notation of order statistics to denote the $x_{(n)}$ as the maximum of $n$ points. This is a very small fraction in high dimensions, showing again that until the end the live points are far from the posterior bulk, and certainly nowhere near the maximum.
\par
The above also implies that the normalised distance between the highest and second highest live point is roughly 
\begin{equation}
    y_{(\nlive)}^{\mathrm{live}} - y_{(\nlive - 1)}^{\mathrm{live}} \approx \frac{2}{d}.
\end{equation}
Before reaching the posterior bulk, $\logLmax - \log\Like^{*} > d/2$, so we must have
\begin{equation}
    \log \Like^{\mathrm{live}}_{(\nlive)} - \log \Like^\mathrm{live}_{(\nlive - 1)} > 1.
\end{equation}
In other words, the highest likelihood point is always at least an order of magnitude greater than the second highest. It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.

\subsubsection*{Aside: nested sampling as a maximiser}
Previous literature \citep{Akrami_2010, Feroz_2011} has explored the potential for nested sampling to be used as a global maximiser, given its ability to handle multi-modalities. In particular, the latter authors emphasised that posterior samplers such as nested sampling find the bulk of the \textit{mass}, not the maximum of the distribution, but that this can be remedied by tightening the termination criterion. We now use the machinery we have developed to put this statement in more quantitative terms. 
\par
A more rigorous derivation (\cref{sec:logLmax}) shows that the maximum live point has mean and variance
\begin{equation}\label{eq:ylivemax}
	\lim_{d,\nlive\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log \nlive}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}.
\end{equation}
Now let us take the dead point to be the termination point with likelihood $\log\Like_\mathrm{end}$ and prior volume $X_\mathrm{end}$, so that
\begin{equation}
	\epsilon = \frac{\int_0^{X_\mathrm{end}} \mathcal{L}\ \mathrm{d}X}{\int_0^\infty \mathcal{L}\ \mathrm{d}X}.
\end{equation}
Note that we have assumed that prior effects are negligible (so $1=\infty$), and that $\epsilon \ll 1$ so that the denominator is approximately the accumulated evidence. Computing this for \eqref{eq:gaussian_logL}, we find the answer in terms of lower incomplete gamma functions
\begin{equation}
\epsilon = 1- \frac{\Gamma_{d/2}\left(X_\mathrm{end}^{2/d}/2\sigma^2\right)}{\Gamma(d/2)}.
\end{equation}
Taking the $X_\mathrm{end}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
\begin{equation}
    \lim_{X_\mathrm{end}\ll (\sqrt{2}\sigma)^d} \epsilon \approx \frac{X_\mathrm{end}}{(\sqrt{2}\sigma)^d \ \Gamma(1+\frac{d}{2})} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{end})^{\frac{d}{2}}}{\Gamma(1+\frac{d}{2})}.
\end{equation}
We thus have an expression relating $\mathcal{L}_\mathrm{end}$ at termination to the termination fraction $\epsilon$. This becomes yet more pleasing in the large $d$ limit, since $\epsilon^{2/d}\to 1$, we find via a Stirling approximation:
\begin{equation}
    \lim_{d\to\infty} \log\mathcal{L}_\mathrm{end} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e}.
\end{equation}
In the event that we keep $\epsilon$ in, we replace $\frac{d}{2e}\to \frac{d}{2e}\epsilon^{2/d}$, so we can of course battle the $\frac{d}{2e}$ term, but this becomes exponentially difficult in high dimensions.
\par
Putting this together, taking $\mathcal{L}_*$ in \eqref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{end}$, and combining this with \eqref{eq:ylivemax} we find
\begin{equation}
    \boxed{
        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
    },
\end{equation}
showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood. The final set of $n$ live points gets you $\log n/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation. 
\par
A plot of the log-likelihood distribution at the end of a run is shown in \cref{fig:logL_distribution}
\begin{figure*}
\begin{center}
    \includegraphics{figures/logL_distribution.pdf}
\end{center}
\caption{Distribution of samples as a function of $\log\mathcal{L}$, showing the posterior $\mathcal{P}(\log\mathcal{L})$, the distribution of the live points $\mathcal{\pi}(\log\mathcal{L} \mid \mathcal{L}>\mathcal{L}_*)$, and the distribution of the maximum likelihood live point $P(\log\mathcal{L}_\mathrm{max}^\mathrm{live})$. The distances are shown between these locations at the end of the run, the key takeaway being that in high dimensions the highest log-likelihood point of a nested sampling run is nowhere near the maximum in high dimensions.}
\label{fig:logL_distribution}
\end{figure*}
